What We Eval in the Shadows

### 2021-08-09 paper/main.tex
Add code generation and give names to sections

%\todo{Say somewhere that the R community commonly calls call/argument capturing
%(with substitute or match.call), Non-Standard Evaluation.}
To understand the how R developers use \eval, we manually inspected top 117 eval
call-sites which contribute to XX\% of the eval calls. We describe the common
usage patterns below.

\subsection{Simplify Interfaces}

A common use of \eval is to \emph{simplify interfaces} by letting users pass
string argument which can be parsed and evaluated in custom a environment. An
example is the \code{rARS} function shown below. This function is defined in the
\code{AdapSamp} package that implements sampling algorithms.

\begin{lstlisting}
rARS <- function(n, formula, min=-Inf, max=Inf, sp) {
    ...
    p <- function(x) { eval(parse(text=formula)) }
    ...
}

\end{lstlisting}
The \code{rARS} function take a \code{formula} argument, an expression in
\code{x} as a string, that represents the kernel of the target distribution.
This is turned into the closure \code{function(x) \{ eval(parse(text=formula))
  \}}. An alternative design would be to supply the closure itself as the
argument, but passing \code{function(x) exp(-x^2/2)} is probably not as
convenient for domain-experts as passing \code{"exp(-x^2/2)"}.

An unexpected use of \eval is the extraction of unevaluated text of arguments
passed to a \code{...} parameter. Consider the definition of the \code{NVL}
function from the \code{statnet.common} package. This function returns the first
argument that is non \code{NULL}.

@@

Assertion packages such as \code{assertthat} provide functions to check pre and
post conditions and produce friendly error messages on failures. These
conditions are written as expressions which are evaluated using \eval. Consider
the \code{see\_if} function shown below. It accepts the conditions as arguments
to the \code{...} parameter. The code of these conditions is extracted using
\code{eval(substitute(alist(...)))}. The conditions are then evaluated
sequentially by the expression \code{eval(assertion, env)}, inside a
\code{tryCatch} block. If a condition evaluates to \code{FALSE}, the function
returns immediately with an error message, leaving the remaining conditions
unevaluated.

@@

Similar idea is used by the \code{igraph} package to define the \code{\%\|\|\%}
operator whose code is shown below. This operator returns the result of
evaluating the first argument (\code{lhs}) if it is not \code{NULL}, otherwise
it returns the result of evaluating the second argument (\code{rhs}). This is
similar to the \code{\|\|} operator which performs the logical OR of its
arguments except it treats \code{NULL} as the \code{FALSE} value.

Many packages use \eval to implement variations of the \code{base} R functions.
For example, the \code{igraph} package provides the \code{do_call} function
shown below. This is a variant of the \code{do.call} function from the
\code{base} package. This function executes a function call by constructs it
from a name and a list of arguments and calling \eval.
\begin{lstlisting}
do_call <- function(f, ..., .args = list(), .env = parent.frame()) {
  f <- substitute(f)
  call <- make_call(f, ..., .args)
  eval(call, .env)
}
\end{lstlisting}

>>>>

%\todo{Say somewhere that the R community commonly calls call/argument capturing
%(with substitute or match.call), Non-Standard Evaluation.}
To understand the how R developers use \eval, we manually inspected top 117 eval
call-sites which contribute to 90\% of the eval calls. We describe the common
usage patterns below.

\emph{Simplify Interfaces}. A common use of \eval is to simplify interfaces by
letting users pass string argument which can be parsed and evaluated in custom a
environment. An example is the \code{rARS} function shown below. This function
is defined in the \code{AdapSamp} package that implements sampling algorithms.

\begin{lstlisting}
rARS <- function(n, formula, min=-Inf, max=Inf, sp) {
    ...
    p <- function(x) { eval(parse(text=formula)) }
    ...
}
\end{lstlisting}

The \code{rARS} function take a \code{formula} argument, an expression in
\code{x} as a string, that represents the kernel of the target distribution.
This is turned into the closure \code{function(x) \{ eval(parse(text=formula))
  \}}. An alternative design would be to supply the closure itself as the
argument, but passing \code{function(x) exp(-x^2/2)} is probably not as
convenient for domain-experts as passing \code{"exp(-x^2/2)"}.

\emph{Code Generation.} The \c{Rcpp} package which bridges R with C++ is a
popular user of this pattern. Its \code{.makeCppMethods} function, shown below,
uses \eval to automatically generate R bindings to native code. The expression
\code{function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))} is
evaluated with different values of \code{what} to generate wrapper functions
which are assigned to the \code{methods} structure keyed by \code{what}.

\begin{lstlisting}
.makeCppMethods <- function(methods, cppclassinfo, env) {
    cppMethods <- names(cppclassinfo@methods)
    newMethods <- names(methods)
    for(what in cppMethods[! cppMethods %in% newMethods])
        methods[[what]] <- eval(substitute(
                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))),
                                env)
    methods
}
\end{lstlisting}

\emph{Extracting ... arguments.} A surprising use of \eval is the extraction of unevaluated text of arguments
passed to a \code{...} parameter. Consider the definition of the \code{NVL}
function from the \code{statnet.common} package. This function returns the first
argument that is non \code{NULL}.

@@

\emph{Conditional Evaluation.} Developers frequently use \eval to explicitly
control the evaluation of code. Consider the \code{\%\|\|\%} operator from the
\code{igraph} package shown below. This operator returns the result of
evaluating the first argument (\code{lhs}) if it is not \code{NULL}, otherwise
it returns the result of evaluating the second argument (\code{rhs}). This is
similar to the \code{\|\|} operator which performs the logical OR of its
arguments except it treats \code{NULL} as the \code{FALSE} value.

A similar idea is used by assertion packages such as \code{assertthat} which
provide functions to check pre and post conditions and produce friendly error
messages on failures. These conditions are written as expressions which are
evaluated using \eval. Consider the \code{see\_if} function shown below. It
accepts the conditions as arguments to the \code{...} parameter. The code of
these conditions is extracted using \code{eval(substitute(alist(...)))}. The
conditions are then evaluated sequentially by the expression
\code{eval(assertion, env)}, inside a \code{tryCatch} block. If a condition
evaluates to \code{FALSE}, the function returns immediately with an error
message, leaving the remaining conditions unevaluated.

@@

\textit{Core function variants.} Many packages use \eval to implement variations of the
\code{base} R functions. For example, the \code{igraph} package provides the
\code{do_call} function shown below. This is a variant of the \code{do.call}
function from the \code{base} package. This function executes a function call by
constructs it from a name and a list of arguments and calling \eval.
\begin{lstlisting}
do_call <- function(f, ..., .args = list(), .env = parent.frame()) {
  f <- substitute(f)
  call <- make_call(f, ..., .args)
  eval(call, .env)
}
\end{lstlisting}
